Sampling requirements to detect ecosystem change

### 2020-04-07 R/manuscript_draft.Rmd
Updates to main manuscript and supp mat before publishing preprint

We detect changepoints with the function _e.divisive_ in the R package _ecp_ [@James2019]. There are several methods available to detect changepoint (reviewed in @James2014); _e.divisive_ is a divisive hierarchical estimation algorithm for multiple change point analysis. We chose this method because it is able to perform multiple change point analysis for both uni- and multi-variate time series, without a priori knowledge of the number of changepoints. Herein, we focus on detecting the most important changepoint (i.e. the one of largest magnitude), although we tested the method on a time-series that would have multiple changepoints (Fig S4). In order to test the performance, we detected the "true" changepoint on the whole time-series, and compare the changepoint found on the sub-sample with the "true" one. The distance to true changepoint served as the performance diagnostic.

@@

```{r simulation_min_time_plot, fig.width = 7.2, fig.height= 3, fig.cap='Minimum number of samples required for 0.8 statistical power given different levels of (a) temporal variability ($\\sigma$), (b) temporal autocorrelation ($\\phi$), and (c) shift size ($\\delta$). Regular sampling was used with the default parameters: $\\sigma = 0.53$, $\\phi = 0.404$, and a shift size $=0.81$ to match the case study. The exact timing of the true changepoint varied for each simulation between time steps 30 and 70. Vertical line indicate respective parameter calculated from the case study time series. The same analyses for random and iterative sampling is in Figs. S2,S3.\\label{fig:simulation_min_time_plot}',message=F,warning=F,echo=F,eval=T}

@@

As another application of our approach, we examined the same sediment core data, but examined total abundance as opposed to community composition (Fig. S5). We tested the three subsampling methods, and it took 11 samples to find the "true" changepoint  (Fig. S6). The initial 5 subsamples analyzed were the same than the subsamples analyzed to answer the question of the change in community  (Fig. \ref{fig:case_study_number_sample}c). <!--However, depending on the metric of interest, the subsampling will target different subsamples to add. In other words, the number of subsamples analyzed can be reduced, especially in cases where the question is well tailored from the beginning.-->

@@

<!---- reexplain and compare the 3 methods (theoretical (Fig 3) vs case study (Fig 4), and how they relate (Fig 2)) ---->
Due to time or funding limitations, there is often a difference between the number of samples collected and the total number of samples that can later be processed. When the processing time is disproportionately higher than the collection time especially, a subsampling can be done prior to processing. A decision must then be made as to which subsamples to analyze. To address this question in the context of detecting changepoints, we tested three subsampling methods: subsampling random points, regular intervals, and an iterative sampling approach (Fig. \ref{fig:conceptual}). We found that the iterative method was systematically better at detecting changes than the two other methods, random subsampling being the least efficient (Figs. \ref{fig:simulation_approaches_plot}, \ref{fig:case_study_number_sample}, S2, S3). Autocorrelation, variance, and shift size, had an effect on how many samples were needed to detect the shift, but did not change which approach was optimal (Fig. \ref{fig:simulation_min_time_plot}).

@@

Multiple subsampling strategies can be chosen (Fig. \ref{fig:conceptual}), but only iterative sampling detected the true changepoint with a limited number of samples (Fig. \ref{fig:simulation_approaches_plot}c). Analyzing 11% of the sample was enough in most cases to approach the "true" changepoint. Applied to the real case study, the iterative method allowed us to find the main changepoint with only 9 samples analyzed (Fig. \ref{fig:case_study_number_sample}). The method also worked well to detect other changepoints of lower magnitude (Fig. S4). @Bruel2018 processed one sample at each centimeter in a 74-centimeter sediment core. Each sample took an average of 3 hours to process. We found that using an iterative approach would have eliminated `r 3*(74-9)` hours of sample processing, or about `r round(3*(74-9)/8)` days, which is just a little over a month of work. This correspond to several thousands of US dollars depending on labor costs.

@@

Our approach goes beyond just paleoecological analyses. Running simulations or using past data to understand the amount of sampling effort required is important in many systems where sample collection or processing is expensive [@White2019e; @White2020]. The specific sampling techniques can also be compared to determine the optimal strategy in terms of accuracy and cost. Our specific approach applies to situation where more subsamples can be added, or processed, after the dynamics occurred [@Zhang2012]. It corresponds very well to the paleoecological approach: samples are taken long after the phenomenon of interest occurred, and allows subsampling at finer or rougher intervals [@Wingard2017]. However, both different types of data and different questions than the one we focused on can be addressed with the same approach. Suppose instead that the goal was to detect a change in relative abundance over time with video-based approaches. It is often not practical to watch entire videos, so it can be useful to choose strategic time-points that would address a specific question of interest. Using an interval sampling approach, one could take a fixed number of samples to start. The trend over time from simple linear regression could be taken. Then, samples can be taken at random locations one-by-one and to see which samples have the largest effect on the trend estimates. If a particular sample has a large effect on the trend, then it would be best to choose another nearby sample. Sampling would continue until the trend estimate reached an equilibrium. Thus, the iterative sampling approach is particularly relevant to data sources where additional samples can be taken long after the initial dynamics. These approaches would also be appropriate for environmental samples, such as water or soil, that can be analyzed later or eDNA that can be extracted from previously-collected samples [@Bohmann2014a].

Our approach is applicable to a wide range of systems and questions, but it does have limitations. When less resources are needed for sample analysis, as opposed to collection, investigators will likely be able to process every sample, and analyzing all samples to obtain a whole picture may be preferred. We note that if resources need to be saved by collecting less samples in the first place, then regular sampling performs better than random sampling (Figs. \ref{fig:simulation_approaches_plot}, S2, S3). Another example where our method present less interest is when addressing questions that require a continuous time series, or at least a regular sampling interval. For example, continuous, high-resolution subsampling of a time-series is generally required to detect critical slowing-down or early warning of shifts [@Frossard2015; @Doncaster2016].

However recent work suggest that combining indicators (in the specific study, trait dynamics and abundance-based early warning signals) allows to forecast population collapse even with at lower resolution and time-series length [@Arkilanian2020]. Extending on the question of being able to detect early warning signals, this objective must be weighed in with the ultimate goal: detecting ecosystem shift. Critical slowing down does not necessarily result in a shift, and a shift can occur without critical slowing down [@Spears2017]. Signs of critical slowing downs are important to understand and recognize because they provide potential early warnings [@Doncaster2016], but in terms of management, knowing the timing of a shift can have larger implication in addressing the driver. Thus, selecting a set number of samples or specific approach may also limit what future questions can be asked. Finally, more subsamples could be processed later to address additional questions.

>>>>

We detect changepoints with the function _e.divisive_ in the R package _ecp_ [@James2019]. There are several methods available to detect changepoint (reviewed in @James2014); _e.divisive_ is a divisive hierarchical estimation algorithm for multiple change point analysis. We chose this method because it is able to perform multiple change point analysis for both uni- and multi-variate time series, without a priori knowledge of the number of changepoints. Herein, we focus on detecting the most important changepoint (i.e. the one of largest magnitude), although we tested the method on a time-series that would have multiple changepoints (Fig. S3). In order to test the performance, we detected the "true" changepoint on the whole time-series, and compare the changepoint found on the sub-sample with the "true" one. The distance to true changepoint served as the performance diagnostic.

@@

```{r simulation_min_time_plot, fig.width = 7.2, fig.height= 3, fig.cap='Minimum number of samples required for 0.8 statistical power given different levels of (a) temporal variability ($\\sigma$), (b) temporal autocorrelation ($\\phi$), and (c) shift size ($\\delta$). Regular sampling was used with the default parameters: $\\sigma = 0.53$, $\\phi = 0.404$, and a shift size $=0.81$ to match the case study. The exact timing of the true changepoint varied for each simulation between time steps 30 and 70. Vertical line indicate respective parameter calculated from the case study time series. The same analyses for random and iterative sampling is in Figs. S1,S2.\\label{fig:simulation_min_time_plot}',message=F,warning=F,echo=F,eval=T}

@@

As another application of our approach, we examined the same sediment core data, but examined total abundance as opposed to community composition (Fig. S4). We tested the three subsampling methods, and it took 11 samples to find the "true" changepoint  (Fig. S5). The initial 5 subsamples analyzed were the same than the subsamples analyzed to answer the question of the change in community  (Fig. \ref{fig:case_study_number_sample}c). <!--However, depending on the metric of interest, the subsampling will target different subsamples to add. In other words, the number of subsamples analyzed can be reduced, especially in cases where the question is well tailored from the beginning.-->

@@

<!---- reexplain and compare the 3 methods (theoretical (Fig 3) vs case study (Fig 4), and how they relate (Fig 2)) ---->
Due to time or funding limitations, there is often a difference between the number of samples collected and the total number of samples that can later be processed. When the processing time is disproportionately higher than the collection time especially, a subsampling can be done prior to processing. A decision must then be made as to which subsamples to analyze. To address this question in the context of detecting changepoints, we tested three subsampling methods: subsampling random points, regular intervals, and an iterative sampling approach (Fig. \ref{fig:conceptual}). We found that the iterative method was systematically better at detecting changes than the two other methods, random subsampling being the least efficient (Figs. \ref{fig:simulation_approaches_plot}, \ref{fig:case_study_number_sample}, S1, S2). Autocorrelation, variance, and shift size, had an effect on how many samples were needed to detect the shift, but did not change which approach was optimal (Fig. \ref{fig:simulation_min_time_plot}).

@@

Multiple subsampling strategies can be chosen (Fig. \ref{fig:conceptual}), but only iterative sampling detected the true changepoint with a limited number of samples (Fig. \ref{fig:simulation_approaches_plot}c). Analyzing 11% of the sample was enough in most cases to approach the "true" changepoint. Applied to the real case study, the iterative method allowed us to find the main changepoint with only 9 samples analyzed (Fig. \ref{fig:case_study_number_sample}). The method also worked well to detect other changepoints of lower magnitude (Fig. S3). @Bruel2018 processed one sample at each centimeter in a 74-centimeter sediment core. Each sample took an average of 3 hours to process. We found that using an iterative approach would have eliminated `r 3*(74-9)` hours of sample processing, or about `r round(3*(74-9)/8)` days, which is just a little over a month of work. This correspond to several thousands of US dollars depending on labor costs.

@@

Our approach goes beyond just paleoecological analyses. Running simulations or using past data to understand the amount of sampling effort required is important in many systems where sample collection or processing is expensive [@White2019e; @White2020]. The specific sampling techniques can also be compared to determine the optimal strategy in terms of accuracy and cost. Our specific approach applies to situations where more subsamples can be added, or processed, after the dynamics occurred [@Zhang2012]. It corresponds very well to paleoecological data: samples are taken long after the phenomenon of interest occurred, and allows subsampling at finer or rougher intervals [@Wingard2017]. However, both different types of data and different questions than those used here can be addressed with the same approach. Suppose instead that the goal was to detect a change in relative abundance over time with video-based approaches. It is often not practical to watch entire videos, so it can be useful to choose strategic time-points that would address a specific question of interest. Using an interval sampling approach, one could take a fixed number of samples to start. The trend over time from simple linear regression could be taken. Then, samples can be taken at random locations one-by-one and to see which samples have the largest effect on the trend estimates. If a particular sample has a large effect on the trend, then it would be best to choose another nearby sample. Sampling would continue until the trend estimate reached an equilibrium. Thus, the iterative sampling approach is particularly relevant to data sources where additional samples can be taken long after the initial dynamics. These approaches would also be appropriate for environmental samples, such as water or soil, that can be analyzed later or eDNA that can be extracted from previously-collected samples [@Bohmann2014a].

Our approach is applicable to a wide range of systems and questions, but it does have limitations. When less resources are needed for sample analysis, as opposed to collection, investigators will likely be able to process every sample, and analyzing all samples to obtain a whole picture may be preferred. We note that if resources need to be saved by collecting less samples in the first place, then regular sampling performs better than random sampling (Figs. \ref{fig:simulation_approaches_plot}, S1, S2). Another example where our method is less useful is when addressing questions that require a continuous time series, or at least a regular sampling interval. For example, continuous, high-resolution subsampling of a time-series is generally required to detect critical slowing-down or early warning of shifts [@Frossard2015; @Doncaster2016].

However, recent work suggest that combining indicators (in the specific study, trait dynamics and abundance-based early warning signals) allows forecasating population collapses even with at lower resolution and time-series length [@Arkilanian2020]. Critical slowing down does not necessarily result in a shift, and a shift can occur without critical slowing down [@Spears2017]. Signs of critical slowing downs are important to understand and recognize because they provide potential early warnings [@Doncaster2016], but in terms of management, knowing the timing of a shift can have larger implications in addressing the underlying driver. Thus, selecting a set number of samples or specific approach may also limit what future questions can be asked.

### 2020-04-07 R/supp_material.Rmd
Updates to main manuscript and supp mat before publishing preprint

source('R/sampling_random.R')
source('R/build_time_series.R')
source('R/updated_param_sensitivity.R')
source('R/calc_min_time2.R')
require(tidyverse)

@@

```{r simulation_min_time_plot_random, fig.width = 7.2, fig.height= 3, fig.cap='Minimum number of samples required for 0.8 statistical power given different levels of (a) temporal variability ($\\sigma$), (b) temporal autocorrelation ($\\phi$), and (c) shift size. Random sampling was used with the default parameters are $\\sigma = 0.53$, $\\phi = 0.404$, and a shift size $=0.81$ to match the case study. The exact timing of the true changepoint varied for each simulation between time steps 30 and 70. Each vertical line indicates the respective parameter calculated from the case study time series. \\label{fig:simulation_min_time_random}',message=F,warning=F,echo=F,eval=T}
require(ggplot2)
require(dplyr)
require(stickylabeller)
par(mfrow=c(1,3))

@@

@@

```{r simulation_min_time_plot_iterative, fig.width = 7.2, fig.height= 3, fig.cap='Minimum number of samples required for 0.8 statistical power given different levels of (a) temporal variability ($\\sigma$), (b) temporal autocorrelation ($\\phi$), and (c) shift size. Iterative sampling was used with the default parameters are $\\sigma = 0.53$, $\\phi = 0.404$, and a shift size $=0.81$ to match the case study. The exact timing of the true changepoint varied for each simulation between time steps 30 and 70. Each vertical line indicates the respective parameter calculated from the case study time series. \\label{fig:simulation_min_time}',message=F,warning=F,echo=F,eval=T}

@@

@@

>>>>

source('R/sampling_random.R')
source('R/build_time_series.R')
source('R/updated_param_sensitivity.R')
source('R/calc_min_time2.R')
require(tidyverse)
require(stickylabeller)

@@

```{r simulation_min_time_plot_random, fig.width = 7.2, fig.height= 3, fig.cap='Minimum number of samples required for 0.8 statistical power given different levels of (a) temporal variability ($\\sigma$), (b) temporal autocorrelation ($\\phi$), and (c) shift size. Random sampling was used with the default parameters are $\\sigma = 0.53$, $\\phi = 0.404$, and a shift size $=0.81$ to match the case study. The exact timing of the true changepoint varied for each simulation between time steps 30 and 70. Each vertical line indicates the respective parameter calculated from the case study time series. \\label{fig:simulation_min_time_random}',message=F,warning=F,echo=F,eval=F}
require(ggplot2)
require(dplyr)
require(stickylabeller)
par(mfrow=c(1,3))

@@

```{r parameter_sensitivity_plot_random,echo=F,message=F,warnings=F,fig.height=4,fig.cap='Random sampling statistical power (fraction of 100 simulations which detected a changepoint within five time points of the true changepoint) for different levels of standard deviation ($\\sigma$), lag-1 autocorrelation ($\\phi$), and shift size ($\\delta$). For each parameter combination, 20 samples were used. An increase in samples would increase the statistical power across this graph.\\label{fig:parameter_sensitivity_plot_random}'}
output_theory_approach <- read.csv(file = 'Output/parameter_sensitivity_random.csv')

require(ggplot2)
require(viridis)
ggplot(aes(x=phi_vec,y=sigma_vec,fill=power),data=output_theory_approach) + geom_tile() +
  scale_fill_viridis(discrete=FALSE,begin = 0.1,end=0.9, limits=c(0,1)) +
  facet_wrap(~shift_size_vec) + 
  theme_classic(base_size = 14, base_family = "")+
  theme(plot.subtitle = element_text(hjust = 0.5)) +
  labs(fill='Power',subtitle = 'Shift size') +
  xlab("Lag-1 autocorrelation (phi)") +
  ylab("Standard deviation (sigma)")
```

@@

```{r simulation_min_time_plot_iterative, fig.width = 7.2, fig.height= 3, fig.cap='Minimum number of samples required for 0.8 statistical power given different levels of (a) temporal variability ($\\sigma$), (b) temporal autocorrelation ($\\phi$), and (c) shift size. Iterative sampling was used with the default parameters are $\\sigma = 0.53$, $\\phi = 0.404$, and a shift size $=0.81$ to match the case study. The exact timing of the true changepoint varied for each simulation between time steps 30 and 70. Each vertical line indicates the respective parameter calculated from the case study time series. \\label{fig:simulation_min_time}',message=F,warning=F,echo=F,eval=F}

@@

```{r parameter_sensitivity_plot_iterative,echo=F,message=F,warnings=F,fig.height=4,fig.cap='Iterative sampling statistical power (fraction of 100 simulations which detected a changepoint within five time points of the true changepoint) for different levels of standard deviation ($\\sigma$), lag-1 autocorrelation ($\\phi$), and shift size ($\\delta$). For each parameter combination, 20 samples were used. An increase in samples would increase the statistical power across this graph.\\label{fig:parameter_sensitivity_plot_iterative}'}
output_theory_approach <- read.csv(file = 'Output/parameter_sensitivity_iterative.csv')

require(ggplot2)
require(viridis)
ggplot(aes(x=phi_vec,y=sigma_vec,fill=power),data=output_theory_approach) + geom_tile() +
  scale_fill_viridis(discrete=FALSE,begin = 0.1,end=0.9) +
  facet_wrap(~shift_size_vec) + 
  theme_classic(base_size = 14, base_family = "")+
  theme(plot.subtitle = element_text(hjust = 0.5)) +
  labs(fill='Power',subtitle = 'Shift size') +
  xlab("Lag-1 autocorrelation (phi)") +
  ylab("Standard deviation (sigma)")
```

@@

require(stickylabeller)
